---
title: "Kidney Project"
author: "Lina, Sherry, Sarah, Demitri, John"
date: "01/05/2020"
output: html_document
---
```{r knitr_options , include=FALSE}
knitr::opts_chunk$set(fig.width=12, fig.height=4,
fig.path='Figs/', warning=FALSE,
message=FALSE, include=FALSE,
echo=FALSE)
set.seed(53079239)
```



```{r}
library(BiocManager)
library(affy)
library(GEOquery)
library(tidyverse)
library(limma)
library(GiANT)
#help(GiANT)
#library(tictoc)
```
# ***To do***:

* Deal with microarray multiple probes for one gene. e.g. there are two rows named ERICH5. We can do this before *renaming the data using annotations* or after *renaming the data using annotations*.

https://www.biostars.org/ provides quite a few solutions to this common issue:

  + https://www.biostars.org/p/47421/

  + https://www.biostars.org/p/74687/

  + https://www.biostars.org/p/336130/

  + https://www.biostars.org/p/74687/

  + https://www.biostars.org/p/396235/

* Subset the Kidney transplant data by the top 50 highly differential common genes (used for identifying Type 2 diabetes) after handling the multiple probes for one gene.

* Run the Random Forest Classifier (best accuracy) for identifying Type 2 diabetes on the Kidney transplant data. Label the samples with predicted outcomes.

* Subset the Kidney transplant samples with predicted Type 2 diabetes. 

* Building the risk calculator using the new data. 

  + Find the highly differential genes e.g. top 50.
  + Build the classifiers.
  + Estimate accuracy and other matrix.

* Shiny App

# Load in data
```{r Load in data}
# takes time
transplant_data <- ReadAffy(celfile.path="GSE14346_RAW")
diabetes_data <- ReadAffy(celfile.path="GSE20966_RAW")

library(GEOquery)
clinical_outcome <-getGEO("GSE120396")
clinical_outcome<- clinical_outcome$GSE120396_series_matrix.txt.gz

print(clinical_outcome$characteristics_ch1.1[1:10])
rejection_status  <- clinical_outcome$characteristics_ch1.1
rejection_status <- unlist(lapply( strsplit(as.character(rejection_status), ": " ) , `[[` , 2)  )
table(rejection_status)
```

```{r}
library(GEOquery)
clinical_outcome <-getGEO("GSE142677")
clinical_outcome<- clinical_outcome$GSE142667_series_matrix.txt.gz
str(clinical_outcome)

print(clinical_outcome$characteristics_ch1.1[1:10])
rejection_status  <- clinical_outcome$characteristics_ch1.1
rejection_status <- unlist(lapply( strsplit(as.character(rejection_status), ": " ) , `[[` , 2)  )
table(rejection_status)
```

# Obtain clinical outcome
```{r Obtain clinical outcome - diabetes}
# diabetes
diabetes_clinical_outcome <- getGEO("GSE20966", GSEMatrix =TRUE, getGPL=FALSE)
if (length(diabetes_clinical_outcome) > 1) idx <- grep("GSE20966", attr(diabetes_clinical_outcome, "names")) else idx <- 1
diabetes_clinical_outcome <- diabetes_clinical_outcome[[idx]]
```

```{r reformat the labels - diabetes}
# diabetes
disease  <- diabetes_clinical_outcome$characteristics_ch1.1
disease <- unlist(lapply( strsplit(as.character(disease), ": " ) , `[[` , 2)  )
table(disease)
```

```{r Obtain clinical outcome - transplant}
#transplant - takes a few seconds to run
transplant_clinical_outcome <- getGEO("GSE14346", GSEMatrix =TRUE, getGPL=FALSE)
if (length(transplant_clinical_outcome) > 1) idx <- grep("GSE14346", attr(transplant_clinical_outcome, "names")) else idx <- 1
transplant_clinical_outcome <- transplant_clinical_outcome[[idx]]

rejection <- transplant_clinical_outcome$`disease state:ch1`
```

```{r reformat the labels - transplant}
rejection_labels <- ifelse(startsWith(rejection, 'STA'),rejection <-"STA",rejection <-"AR") 

table(rejection_labels)
```
STA are stable patients and AR are patients with acute rejection.

# Pre-processing
```{r Obtaining scaled expression values - diabetes}
eset_diabetes <- rma(diabetes_data)
```
```{r Obtaining scaled expression values - transplant}
eset_transplant = rma(transplant_data)
```
# Data in table format
```{r}
diabetes = exprs(eset_diabetes)
```

```{r}
transplant = exprs(eset_transplant)
```
# →To do: Handle multiple probes for one gene - we can do this before or after renaming the data using annotations

# Rename the data using annotations
#### Diabetes
```{r diabetes data}
library("annotate")
annotation(diabetes_data)
```
```{r}
#BiocManager::install("u133x3p.db")
```

```{r}
library("u133x3p.db")
diabetes_genenames <- as.character(rownames(diabetes))
diabetes_sym_uncleaned <- getSYMBOL(diabetes_genenames, "u133x3p")
```

```{r}
# remove NA
diabetes_sym = na.omit(diabetes_sym_uncleaned)

# filter data
diabetes_filtered = diabetes[which(rownames(diabetes) %in% names(diabetes_sym)), ]

# rename genes
rownames(diabetes_filtered) <- diabetes_sym
```

#### Transplant
```{r}
library("annotate")
annotation(transplant_data)
```
```{r}
#BiocManager::install("hgu133plus2.db")
```

```{r}
library("hgu133plus2.db")
transplant_genenames <- as.character(rownames(transplant))
transplant_sym_uncleaned <- getSYMBOL(transplant_genenames, "hgu133plus2")
```

```{r}
# remove NA
transplant_sym = na.omit(transplant_sym_uncleaned)

# filter data
transplant_filtered = transplant[which(rownames(transplant) %in% names(transplant_sym)), ]

# rename genes
rownames(transplant_filtered) <- transplant_sym
```


# Multiple probes for one gene

https://rdrr.io/cran/GiANT/man/mergeProbesForGenes.html

```{r}
nrow(transplant_filtered)
#unique(rownames(transplant_filtered))
nrow(diabetes_filtered)
```

```{r}
transplant_filtered2 <- mergeProbesForGenes(transplant_filtered, method = "max")
diabetes_filtered2 <- mergeProbesForGenes(diabetes_filtered, method = "max")
dim(transplant_filtered2)
dim(diabetes_filtered2)

```


# Common genes
```{r}
common_genes = rownames(diabetes_filtered2)[rownames(diabetes_filtered2) %in% rownames(transplant_filtered2)]
length(common_genes)
```

# Subset diabetes data - only keep common genes as transplant data
```{r}
diabetes_with_com_genes = diabetes_filtered2[which(rownames(diabetes_filtered2) %in% common_genes), ]
dim(diabetes_with_com_genes)
```


# Subset transplant data - only keep common genes as diabetes data
```{r}
transplant_with_com_genes = transplant_filtered2[which(rownames(transplant_filtered2) %in% common_genes), ]
dim(transplant_with_com_genes)
```



# Highly differential genes - Type 2 diabetes
```{r set-up}
# encode yes and no to 1 and 0 
disease_encode <- ifelse(disease == "non-diabetic control", 0, 1 )
# set group names
groupname <- paste("G", disease_encode, sep="")    

design <- model.matrix(~ groupname + 0)
```

```{r limma}
fit <- lmFit(diabetes_with_com_genes, design)
cont.matrix <- makeContrasts(groupnameG0-groupnameG1, levels=design)
fit2 <- contrasts.fit(fit, cont.matrix)
fit2 <- eBayes(fit2)

table(fit2$p.value <= 0.01) # number of insignificant and significant genes
```

# Subset the data by the significant genes
This could improve accuracy of classifiers. We can select the top 50, 150, 250 etc. 

```{r top 30 genes}
tt_top10 <- topTable(fit2,n = 10)

genecode_top10 <- as.character(rownames(tt_top10)) # top 50 genes by gene code

#genenames_top50 <- as.character(tt_top50$Id)# top 50 genes by gene names

#gse_8f = gse[which(rownames(gse) %in% genenames_top8), ]
#table(rownames(diabetes_with_com_genes) %in% genecode_top50)


diabetes_top10_genes = diabetes_with_com_genes[which(rownames(diabetes_with_com_genes) %in% genecode_top10), ]
dim(diabetes_top10_genes)
```

```{r top 50 genes}
tt_top50 <- topTable(fit2,n = 50)

genecode_top50 <- as.character(rownames(tt_top50)) # top 50 genes by gene code

#genenames_top50 <- as.character(tt_top50$Id)# top 50 genes by gene names

#gse_8f = gse[which(rownames(gse) %in% genenames_top8), ]
#table(rownames(diabetes_with_com_genes) %in% genecode_top50)


diabetes_top50_genes = diabetes_with_com_genes[which(rownames(diabetes_with_com_genes) %in% genecode_top50), ]
dim(diabetes_top50_genes)
```

```{r top 150 genes}
tt_top150 <- topTable(fit2,n = 150)

genecode_top150 <- as.character(rownames(tt_top150)) # top 150 genes by gene code

#genenames_top150 <- as.character(tt_top150$ID)# top 150 genes by gene names

diabetes_top150_genes = diabetes_with_com_genes[which(rownames(diabetes_with_com_genes) %in% genecode_top150), ]
dim(diabetes_top150_genes)
```

```{r top 250 genes}
tt_top250 <- topTable(fit2,n = 250)

genecode_top250 <- as.character(rownames(tt_top250)) # top 250 genes by gene code

#genenames_top250 <- as.character(tt_top250$ID)# top 250 genes by gene names

diabetes_top250_genes = diabetes_with_com_genes[which(rownames(diabetes_with_com_genes) %in% genecode_top250), ]
dim(diabetes_top250_genes)
```


# Type 2 Diabetes Classifier 
### On diabetes_with_com_genes (need around 100 seconds to run.)
Note: I didn't subset the data. The data used here contains both the significant and insignificant genes.
```{r}
## quick filter to reduce computational time
largevar = apply(diabetes_with_com_genes, 1, var)
ind = which(largevar > quantile(largevar, 0.9))

X = as.matrix(t(diabetes_with_com_genes[ind,]))
y = disease

cvK = 5  # number of CV folds
cv_50acc5_knn = cv_50acc5_svm = cv_50acc5_rf = c()
cv_acc_knn = cv_acc_svm = cv_acc_rf = c()

n_sim = 25 ## number of repeats
for (i in 1:n_sim) {

  cvSets = cvTools::cvFolds(nrow(X), cvK)  # permute all the data, into 5 folds
  cv_acc_knn = cv_acc_svm = cv_acc_rf = c()
  
  for (j in 1:cvK) {
    test_id = cvSets$subsets[cvSets$which == j]
    X_test = X[test_id, ]
    X_train = X[-test_id, ]
    y_test = y[test_id]
    y_train = y[-test_id]
    
    ## KNN
    fit5 = class::knn(train = X_train, test = X_test, cl = y_train, k = 5)
    cv_acc_knn[j] = table(fit5, y_test) %>% diag %>% sum %>% `/`(length(y_test))
    
    ## SVM
    svm_res <- e1071::svm(x = X_train, y = as.factor(y_train))
    fit <- predict(svm_res, X_test)
    cv_acc_svm[j] = table(fit, y_test) %>% diag %>% sum %>% `/`(length(y_test))

    ## RandomForest
    rf_res <- randomForest::randomForest(x = X_train, y = as.factor(y_train))
    fit <- predict(rf_res, X_test)
    cv_acc_rf[j] = table(fit, y_test) %>% diag %>% sum %>% `/`(length(y_test))
  }
  cv_50acc5_knn <- append(cv_50acc5_knn, mean(cv_acc_knn))
  cv_50acc5_svm <- append(cv_50acc5_svm, mean(cv_acc_svm))
  cv_50acc5_rf <- append(cv_50acc5_rf, mean(cv_acc_rf))
} ## end for
```

```{r}
boxplot(list(SVM = cv_50acc5_svm, KNN = cv_50acc5_knn , RF= cv_50acc5_rf), main = "CV Accuracy")
```

### On top 50 genes
```{r}
## quick filter to reduce computational time
largevar = apply(diabetes_top50_genes, 1, var)
ind = which(largevar > quantile(largevar, 0.9))

X = as.matrix(t(diabetes_top50_genes[ind,]))
y = disease

cvK = 5  # number of CV folds
cv_50acc5_knn = cv_50acc5_svm = cv_50acc5_rf = c()
cv_acc_knn = cv_acc_svm = cv_acc_rf = c()

n_sim = 25 ## number of repeats
for (i in 1:n_sim) {

  cvSets = cvTools::cvFolds(nrow(X), cvK)  # permute all the data, into 5 folds
  cv_acc_knn = cv_acc_svm = cv_acc_rf = c()
  
  for (j in 1:cvK) {
    test_id = cvSets$subsets[cvSets$which == j]
    X_test = X[test_id, ]
    X_train = X[-test_id, ]
    y_test = y[test_id]
    y_train = y[-test_id]
    
    ## KNN
    fit5 = class::knn(train = X_train, test = X_test, cl = y_train, k = 5)
    cv_acc_knn[j] = table(fit5, y_test) %>% diag %>% sum %>% `/`(length(y_test))
    
    ## SVM
    svm_res <- e1071::svm(x = X_train, y = as.factor(y_train))
    fit <- predict(svm_res, X_test)
    cv_acc_svm[j] = table(fit, y_test) %>% diag %>% sum %>% `/`(length(y_test))

    ## RandomForest
    rf_res <- randomForest::randomForest(x = X_train, y = as.factor(y_train))
    fit <- predict(rf_res, X_test)
    cv_acc_rf[j] = table(fit, y_test) %>% diag %>% sum %>% `/`(length(y_test))
  }
  cv_50acc5_knn <- append(cv_50acc5_knn, mean(cv_acc_knn))
  cv_50acc5_svm <- append(cv_50acc5_svm, mean(cv_acc_svm))
  cv_50acc5_rf <- append(cv_50acc5_rf, mean(cv_acc_rf))
} ## end for
```

```{r}
boxplot(list(SVM = cv_50acc5_svm, KNN = cv_50acc5_knn , RF= cv_50acc5_rf), main = "Top 50 genes - CV Accuracy")
```

### On top 150 genes
```{r}
## quick filter to reduce computational time
largevar = apply(diabetes_top150_genes, 1, var)
ind = which(largevar > quantile(largevar, 0.9))

X = as.matrix(t(diabetes_top150_genes[ind,]))
y = disease

cvK = 5  # number of CV folds
cv_50acc5_knn = cv_50acc5_svm = cv_50acc5_rf = c()
cv_acc_knn = cv_acc_svm = cv_acc_rf = c()

n_sim = 25 ## number of repeats
for (i in 1:n_sim) {

  cvSets = cvTools::cvFolds(nrow(X), cvK)  # permute all the data, into 5 folds
  cv_acc_knn = cv_acc_svm = cv_acc_rf = c()
  
  for (j in 1:cvK) {
    test_id = cvSets$subsets[cvSets$which == j]
    X_test = X[test_id, ]
    X_train = X[-test_id, ]
    y_test = y[test_id]
    y_train = y[-test_id]
    
    ## KNN
    fit5 = class::knn(train = X_train, test = X_test, cl = y_train, k = 5)
    cv_acc_knn[j] = table(fit5, y_test) %>% diag %>% sum %>% `/`(length(y_test))
    
    ## SVM
    svm_res <- e1071::svm(x = X_train, y = as.factor(y_train))
    fit <- predict(svm_res, X_test)
    cv_acc_svm[j] = table(fit, y_test) %>% diag %>% sum %>% `/`(length(y_test))

    ## RandomForest
    rf_res <- randomForest::randomForest(x = X_train, y = as.factor(y_train))
    fit <- predict(rf_res, X_test)
    cv_acc_rf[j] = table(fit, y_test) %>% diag %>% sum %>% `/`(length(y_test))
  }
  cv_50acc5_knn <- append(cv_50acc5_knn, mean(cv_acc_knn))
  cv_50acc5_svm <- append(cv_50acc5_svm, mean(cv_acc_svm))
  cv_50acc5_rf <- append(cv_50acc5_rf, mean(cv_acc_rf))
} ## end for
```

```{r}
boxplot(list(SVM = cv_50acc5_svm, KNN = cv_50acc5_knn , RF= cv_50acc5_rf), main = "Top 150 genes - CV Accuracy")
```

### On top 250 genes
```{r}
## quick filter to reduce computational time
largevar = apply(diabetes_top250_genes, 1, var)
ind = which(largevar > quantile(largevar, 0.9))

X = as.matrix(t(diabetes_top250_genes[ind,]))
y = disease

cvK = 5  # number of CV folds
cv_50acc5_knn = cv_50acc5_svm = cv_50acc5_rf = c()
cv_acc_knn = cv_acc_svm = cv_acc_rf = c()

n_sim = 25 ## number of repeats
for (i in 1:n_sim) {

  cvSets = cvTools::cvFolds(nrow(X), cvK)  # permute all the data, into 5 folds
  cv_acc_knn = cv_acc_svm = cv_acc_rf = c()
  
  for (j in 1:cvK) {
    test_id = cvSets$subsets[cvSets$which == j]
    X_test = X[test_id, ]
    X_train = X[-test_id, ]
    y_test = y[test_id]
    y_train = y[-test_id]
    
    ## KNN
    fit5 = class::knn(train = X_train, test = X_test, cl = y_train, k = 5)
    cv_acc_knn[j] = table(fit5, y_test) %>% diag %>% sum %>% `/`(length(y_test))
    
    ## SVM
    svm_res <- e1071::svm(x = X_train, y = as.factor(y_train))
    fit <- predict(svm_res, X_test)
    cv_acc_svm[j] = table(fit, y_test) %>% diag %>% sum %>% `/`(length(y_test))

    ## RandomForest
    rf_res <- randomForest::randomForest(x = X_train, y = as.factor(y_train))
    fit <- predict(rf_res, X_test)
    cv_acc_rf[j] = table(fit, y_test) %>% diag %>% sum %>% `/`(length(y_test))
  }
  cv_50acc5_knn <- append(cv_50acc5_knn, mean(cv_acc_knn))
  cv_50acc5_svm <- append(cv_50acc5_svm, mean(cv_acc_svm))
  cv_50acc5_rf <- append(cv_50acc5_rf, mean(cv_acc_rf))
} ## end for
```

```{r}
boxplot(list(SVM = cv_50acc5_svm, KNN = cv_50acc5_knn , RF= cv_50acc5_rf), main = "Top 250 genes - CV Accuracy")
```

# →To do: prepare data of top 10 genes for both diabetes and transplant data

```{r}
diabetes_top10_genes = diabetes_with_com_genes[which(rownames(diabetes_with_com_genes) %in% genecode_top10), ]
dim(diabetes_top10_genes)
transplant_top10_genes = transplant_with_com_genes[which(rownames(transplant_with_com_genes) %in% genecode_top10),]

dim(transplant_top10_genes)
```

```{r}
diabetes_top50_genes = diabetes_with_com_genes[which(rownames(diabetes_with_com_genes) %in% genecode_top50), ]
dim(diabetes_top50_genes)
transplant_top50_genes = transplant_with_com_genes[which(rownames(transplant_with_com_genes) %in% genecode_top50),]

dim(transplant_top50_genes)
```
```{r}
diabetes_top150_genes = diabetes_with_com_genes[which(rownames(diabetes_with_com_genes) %in% genecode_top150), ]
dim(diabetes_top150_genes)
transplant_top150_genes = transplant_with_com_genes[which(rownames(transplant_with_com_genes) %in% genecode_top150),]

dim(transplant_top150_genes)
```

# →To do: Append predicted Type 2 Diabetes labels to the Kidney Transplant data samples

Since classifier with top 50 genes has the highest accuracy, we choose this to classify the kidney transplant data(with or with our diabetes)

```{r}
#Need to renamethe classifiers or run this chunck first for the top 10 genes classifier

## quick filter to reduce computational time
#largevar = apply(diabetes_top50_genes, 1, var)
#ind = which(largevar > quantile(largevar, 0.8))

set.seed(0123456)
X = as.matrix(t(diabetes_top10_genes))
y = disease

cvK = 3  # number of CV folds
cv_50acc5_knn = cv_50acc5_svm = cv_50acc5_rf = c()
cv_acc_knn = cv_acc_svm = cv_acc_rf = c()

n_sim = 150 ## number of repeats
for (i in 1:n_sim) {

  cvSets = cvTools::cvFolds(nrow(X), cvK)  # permute all the data, into 5 folds
  cv_acc_knn = cv_acc_svm = cv_acc_rf = c()
  
  for (j in 1:cvK) {
    test_id = cvSets$subsets[cvSets$which == j]
    X_test = X[test_id, ]
    X_train = X[-test_id, ]
    y_test = y[test_id]
    y_train = y[-test_id]
    
    ## KNN
    fit5 = class::knn(train = X_train, test = X_test, cl = y_train, k = 5)
    cv_acc_knn[j] = table(fit5, y_test) %>% diag %>% sum %>% `/`(length(y_test))
    
    ## SVM
    svm_res <- e1071::svm(x = X_train, y = as.factor(y_train))
    fitsvm <- predict(svm_res, X_test)
    cv_acc_svm[j] = table(fitsvm, y_test) %>% diag %>% sum %>% `/`(length(y_test))

    ## RandomForest
    rf_res <- randomForest::randomForest(x = X_train, y = as.factor(y_train))
    fitrf <- predict(rf_res, X_test)
    cv_acc_rf[j] = table(fitrf, y_test) %>% diag %>% sum %>% `/`(length(y_test))
  }
  cv_50acc5_knn <- append(cv_50acc5_knn, mean(cv_acc_knn))
  cv_50acc5_svm <- append(cv_50acc5_svm, mean(cv_acc_svm))
  cv_50acc5_rf <- append(cv_50acc5_rf, mean(cv_acc_rf))
} ## end for
```



```{r}
boxplot(list(KNN = cv_50acc5_knn, SVM =cv_50acc5_svm,  RF = cv_50acc5_rf), xlab = "Classifier", ylab = "Accuracy", main = "Accuracy of the Type 2 diabetes classifiers", sub="Performance comparision of classifiers: Accuracy scores of 3-fold 150 repeats Cross-Validation", cex.main=1.3, cex.lab=1.1, cex.axis=1, cex.sub = 0.8, col=c("chocolate1","chartreuse3","cyan3","darkgoldenrod1"), border=c ("chocolate4","chartreuse4","cyan4","darkgoldenrod4"))
```

```{r}
boxplot(list(SVM = cv_50acc5_svm, KNN = cv_50acc5_knn , RF= cv_50acc5_rf), main = "Top 10 genes - CV Accuracy")
```

```{r}
X_train[2,]
diabetes_top_10_genes
var=rownames(diabetes_top_10_genes)
transplant_top10_genes_var = transplant_with_com_genes[which(rownames(transplant_with_com_genes) %in% var),]
#saveRDS(transplant_top10_genes_var, "transplant_top10_genes.rds")
dim(X)
dim(transplant_top10_genes_var)
```

```{r}
#convert into correct format 
test<-as.matrix(t(transplant_top10_genes))
dim(test)
```


```{r}
#predict transplant data using rf method
#write.csv(test[1,],"diabetes_input1")
di=read.csv("diabetes_input1")

X_train[2,]
t=t(as.matrix(di))[2,]

list<-c(1,2,3,1,1,1,1,1,1,1)
new=t(as.matrix(list))

p=predict(rf_res, X_train[10,]) 
pp <- ifelse(p == "non-diabetic control", "No diabetes", "Yes diabetes" )


print(pp)
```

```{r}
#predict transplant data using svm method

p_svm<-predict(svm_res,newdata = test)
table(ifelse(p_svm == "non-diabetic control", "No diabetes", "Yes diabetes" ))

```




# →To do: Subset the Kidney transplant samples with predicted Type 2 diabetes
Note we are using the prediction from RF method (with top 10 selected features)

```{r}
#rename the colnames with the prediction from our classifier (easy identification of diabetes or not)
colnames(transplant_with_com_genes)= pp
#head(transplant_with_com_genes)
dim(transplant_with_com_genes)
```

```{r}
#filter transplant data - only keeping the patients with diabetes
t_transplant_with_com_genes=t(transplant_with_com_genes)
transplant_filter_diabetes = t_transplant_with_com_genes[which(rownames(t_transplant_with_com_genes) =="Yes diabetes"),]

transplant_filter_diabetes=t(transplant_filter_diabetes)
dim(transplant_filter_diabetes)
#-starts_with("Petal")
```

#  →To do: Highly differential genes - Rejection vs non-rejection

```{r }

# encode yes and no to 1 and 0 
rejection_encode <- ifelse(rejection_labels == "STA", 0, 1 )
TF=rownames(t_transplant_with_com_genes) =="Yes diabetes"

rejection_encode_filtered=rejection_encode[TF==TRUE]

#filter rejection labels only for those that are predicted diabetes

# set group names
groupname <- paste("G", rejection_encode_filtered, sep="")    

design <- model.matrix(~ groupname + 0)
table(rejection_encode_filtered)
```

```{r}
fit_transplant <- lmFit(transplant_filter_diabetes, design)
cont.matrix <- makeContrasts(groupnameG0-groupnameG1, levels=design)
fit_transplant2 <- contrasts.fit(fit_transplant, cont.matrix)
fit_transplant2 <- eBayes(fit_transplant2)
#help(eBayes)
tT <- topTable(fit_transplant2,number =50)
table(fit_transplant2$p.value <= 0.0005) 
```



# →To do: Subset the transplant data by highly differential genes - Rejection vs non-rejection


selecting top 50 genes, we can test different numbers too.

```{r top 10 genes}
tT <- topTable(fit_transplant2,n = 10)

transplant_genecode_top10 <- as.character(rownames(tT)) # top 50 genes by gene code

#transplant_genenames_top10 <- as.character(tT$ID)# top 50 genes by gene names

transplant_10_genes = transplant_filter_diabetes[which(rownames(transplant_filter_diabetes) %in% transplant_genecode_top10), ]

dim(transplant_10_genes)
unique(rownames(transplant_10_genes))
# note the multiple probes with same genes here... but i'll continue for now
```




# →To do: Classifier - Rejection vs. non-rejection 

For now simply a draft of predicting rejection and non-rejection. note that we dont have any samples with diabetes


```{r}

## quick filter to reduce computational time

#largevar = apply(gse, 1, var)
#ind = which(largevar > quantile(largevar, 0.9))
set.seed(0123456)
#library(glmnet)
X = as.matrix(t(transplant_10_genes))
y = rejection_encode_filtered

cvK = 5  # number of CV folds
cv_50acc5_knn = cv_50acc5_svm = cv_50acc5_rf =cv_50eval_cox = c()
cv_acc_knn = cv_acc_svm = cv_acc_rf =cv_acc_lasso= c()

n_sim = 25 ## number of repeats
for (i in 1:n_sim) {

  cvSets = cvTools::cvFolds(nrow(X), cvK)  # permute all the data, into 5 folds
  cv_acc_knn = cv_acc_svm = cv_acc_rf = c()
  
  for (j in 1:cvK) {
    test_id = cvSets$subsets[cvSets$which == j]
    X_test = X[test_id, ]
    X_train = X[-test_id, ]
    y_test = y[test_id]
    y_train = y[-test_id]
    
    ## KNN
    fit5 = class::knn(train = X_train, test = X_test, cl = y_train, k = 5)
    cv_acc_knn[j] = table(fit5, y_test) %>% diag %>% sum %>% `/`(length(y_test))
    
    ## SVM
    svm_res <- e1071::svm(x = X_train, y = as.factor(y_train))
    fit <- predict(svm_res, X_test)
    cv_acc_svm[j] = table(fit, y_test) %>% diag %>% sum %>% `/`(length(y_test))

    ## RandomForest
    rf_res <- randomForest::randomForest(x = X_train, y = as.factor(y_train))
    fit <- predict(rf_res, X_test)
    cv_acc_rf[j] = table(fit, y_test) %>% diag %>% sum %>% `/`(length(y_test))
    
    
  }
  cv_50acc5_knn <- append(cv_50acc5_knn, mean(cv_acc_knn))
  cv_50acc5_svm <- append(cv_50acc5_svm, mean(cv_acc_svm))
  cv_50acc5_rf <- append(cv_50acc5_rf, mean(cv_acc_rf))
  
} ## end for


```
```{r}
B1=boxplot(list(SVM = cv_50acc5_svm, KNN = cv_50acc5_knn , RF= cv_50acc5_rf))

```

Note that we will need to calculate the overall accuracy (for the 2 level classification)


Graphics to include in shiny app

How many diabetes patients get rejection and where you are- stacked bar graph
Ask fro female or male
The bar graph can have female and male output

```{r}
#number of patients (predicted with) diabetes that have rejection / no rejections 
t=as.data.frame(as.matrix(table(rejection_encode_filtered)))

d=as.matrix(t)
#d %>% mutate(status=rejection)
data=as.data.frame(d)

data$rejection_status=ifelse(rownames(data)=="0","no rejection","rejection")



bp<-ggplot(data, aes(fill=rejection_status,y=V1,x="")) + geom_bar(width =1, stat="identity") +ggtitle("Rejection status for patients predicted with Type 2 Diabetes")+ labs(y = "Percentage",x="Diabetes patients(predicted)")+geom_text(aes(label = paste(round(V1 / sum(V1) * 100, 1), "%")),
            position = position_stack(vjust = 0.5))+
  theme_classic() +
  theme(plot.title = element_text(hjust=0.5),
        axis.line = element_blank(),
        axis.text = element_blank(),
        axis.ticks = element_blank()) 
bp
```

```{r}
#create theme
blank_theme <- theme_minimal()+
  theme(
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  panel.border = element_blank(),
  panel.grid=element_blank(),
  axis.ticks = element_blank(),
  plot.title=element_text(size=14, face="bold")
  )
```

```{r}
library(ggrepel)
pie <- bp + coord_polar("y", start=0) + scale_fill_brewer(palette="Dark2")+ blank_theme +
  theme(axis.text.x=element_blank()) +labs(fill = "Rejection Status")
#+geom_label_repel(aes(label = status), size=5, show.legend = F, nudge_x = 1)
pie
```


```{r}
#p=predict(svm_res, test) 
#pp <- ifelse(p == "non-diabetic control", "No diabetes", "Yes diabetes" )


#table(pp)
```

```{r}
diabetes_top_10_genes<-readRDS("diabetes_top10_genes.rds")
rejection_encode_filtered<-read_rds("rejection_encode_filtered.rds")
transplant_top10_genes<-readRDS("transplant_top10_genes.rds")
disease<-readRDS("disease.rds")
test<-readRDS("test.rds")
```

```{r}

 
genemeans = t(apply(transplant_top10_genes[,-1], 1, mean))

inputs=as.numeric(c(1,2,3,4,5,6,7,8,9,10))

#ggplot(genemeans, aes( x=genemeans[,1],y = transplant_top10_genes$mean))
#plot(genemeans, main="Genes comparison",xlab="genes ", ylab="Gene values ", pch=19)
```


```{r}
genemeans = t(apply(transplant_top10_genes[,-1], 1, mean))
df <- data.frame(gene_names = rownames(transplant_top10_genes),
                 values = as.numeric(genemeans),
                inputvalues=inputs )

colors <- c("Other patients" = "cornflowerblue", "You"="red")
library(ggpubr)
ggplot(df, aes(x = gene_names))+ 
  geom_point(aes(x = gene_names,y = values,color = "Other patients"),size = 1.5) +
  geom_point(aes(x = gene_names,y = inputvalues,color = "You" , shape =17),size = 1.5)+
  scale_shape_manual(values=c(3, 16, 17))+
  scale_shape_identity()+
  ggtitle("Gene comparisons")+
  theme(plot.title = element_text(hjust = 0.5))+  
  theme(
    legend.position = c(0.95, 0.95),
    legend.justification = c("right", "top"),
    legend.title = element_text( size = 12)
  )+ labs(color='Legend')  +xlab("Genes") + ylab("Values")



```


```{r}
df
```


```{r}
load("classifiers.RData")
```


```{r}
Classifiers.RData
```

